---
title: "MetaCell based analyses - Tumor compartment"
author: "Kaspar Bresser"
date: "20/10/2021"
output: 
  github_document:
    toc: true
  html_document: 
    theme: simplex
    highlight: pygments
#    code_folding: show
    self_contained: TRUE
    toc: yes
    toc_float:
      collapsed: no
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      autodep  = TRUE,
                      cache = FALSE,
                      fig.width = 5,
                      fig.asp = 0.618,
                      fig.align = "center")
```


Used the analysis below to investigate genes of interest in Tumor MetaCell 12, which was specifically enriched in QPCTL deficient TMEs


```{r loading, echo=FALSE}
library(here)
library(Seurat)
library(fgsea)
library(metacell)
library(msigdbr)
library(ggpubr)
library(tidyverse)

```


## Import data

Get the metacell object

```{r import}
scdb_init(here("Data", "Metacell_files_tumor"))
mc <- scdb_mc("QPCTL_Tum_MC2")
```

## Plot marker genes

Check the top and bottom 5 genes of MC12. 

```{r get_markers}
lfp <- as.data.frame(log2(mc@mc_fp))

lfp %>% 
  rownames_to_column("genes") %>% 
  dplyr::select("genes", "12") %>% 
  filter(dense_rank(`12`) <= 5 | dense_rank(-`12`) <= 5) -> TopBot.MC12

TopBot.MC12
```


and plot as waterfall.

```{r top_bottom_plot, fig.asp=1.2, fig.width=3}
TopBot.MC12 %>% 
  mutate(genes = reorder(genes, `12`)) %>% 
  ggplot(aes(x = genes, y = `12`, fill = `12` > 0))+
    geom_bar(stat = "identity")+
    scale_fill_manual(values=c("blue", "red"))+
    labs(title = "top and bottom MC12", x = "gene", y = "log2 enrichment")+
    theme(legend.position = "none", plot.title = element_text(hjust = -1))+
    coord_flip()

ggsave(here("Figs", "tumor", "MC12_top_bottom.pdf"), width = 1, height = 2, scale = 2.5)
```

## GSEA MC12

First get the top and bottom 200 genes of MC12. Will use as input for fgsea analysis.

```{r get_stats}
lfp %>% 
  rownames_to_column("genes") %>% 
  dplyr::select("genes", "12") %>% 
  deframe() %>% 
  sort(decreasing = T) -> stats

# subset to 200 highest and 200 lowest values
stats <- stats[dense_rank(stats) <= 200 | dense_rank(-stats) <= 200]

str(stats)
```


Now import the pathways

```{r get_pathways}
pathways.hallmark <- as.data.frame(msigdbr(species = "Mus musculus", category = "H"))
pathways <- split(pathways.hallmark$gene_symbol, pathways.hallmark$gs_name)

pathways %>% 
  head(10) %>% 
  str()
```

And perform the analysis. 

```{r run_fgsea}
fgseaRes <- fgsea(pathways=pathways, stats=stats, minSize = 10)
```


Now we can plot the enrichment of the pathways foudn by fgsea. 

```{r plot_enrichment, fig.asp=.8}
fgseaRes %>% 
  filter(pval < 0.05)  %>% 
  mutate(pathway = gsub("HALLMARK_", "", pathway),
         pathway = gsub("_", " ", pathway),
         pathway = reorder(pathway, NES)) %>% 
  ggplot( aes(x = NES, y = pathway))+
    geom_point(aes(size = size, color  = padj))+
    grids(axis = "y", linetype = "dashed")+
    geom_vline(xintercept = 0, color = "blue", linetype = "dotted")

ggsave(filename = here("Figs", "tumor", "MC12_Pathways_Hallmark.pdf"), device = "pdf", width = 4.5, height = 4, useDingbats=FALSE )

fgseaRes %>% 
  filter(pval < 0.05)  %>% 
  write_tsv(here("Output", "tumor_MC12_GSEA_Hallmark.tsv"))
```


And plot the top scoring pathways. 

```{r example_1}
pw <- "HALLMARK_INTERFERON_GAMMA_RESPONSE"

fgseaRes %>%  
  dplyr::filter(pathway == pw) %>%  
  pull(NES) %>% round(digits = 2) -> nes

plotEnrichment(pathway = pathways[[pw]], stats)+
  ggtitle(pw)+
  geom_text(aes(label = paste0(c("NES = ", as.character(nes)), collapse = ""), x = Inf, y = Inf  ), 
            vjust = "inward", hjust = "inward")



ggsave(filename = here("Figs", "tumor", "MC12_Pathway_IFNy.pdf"), device = "pdf", width = 5, height = 3, useDingbats=FALSE )
```

```{r example_2}
pw <- "HALLMARK_INTERFERON_ALPHA_RESPONSE"

fgseaRes %>%  
  dplyr::filter(pathway == pw) %>%  
  pull(NES) %>% round(digits = 2) -> nes

plotEnrichment(pathway = pathways[[pw]], stats)+
  ggtitle(pw)+
  geom_text(aes(label = paste0(c("NES = ", as.character(nes)), collapse = ""), x = Inf, y = Inf  ), 
            vjust = "inward", hjust = "inward")



ggsave(filename = here("Figs", "tumor", "MC12_Pathway_IFNa.pdf"), device = "pdf", width = 5, height = 3, useDingbats=FALSE )
```


## Violin plots

Next, we'll check how various IFN responsive genes are expressed across the different MCs.

Import the seurat object.

```{r import_seurat}
seurat.object <- read_rds(here("Data", "seurat_Tum.rds"))
```

Normalize and scale the data.

```{r normalize}
seurat.object <- NormalizeData(seurat.object, assay = "RNA", normalization.method = "CLR")
seurat.object <- ScaleData(seurat.object, assay = "RNA")
```


Extract the expression data for all genes in the IFNg geneset.

```{r get_expression_IFN}
IFN.genes <- pathways[["HALLMARK_INTERFERON_GAMMA_RESPONSE"]]

seurat.object %>% 
  GetAssayData( slot = "scale.data", assay = "RNA")%>% 
  as.matrix() %>%
  t() %>% 
  as.data.frame() %>% 
  as_tibble(rownames = "cellcode") %>% 
  dplyr::select(one_of(c("cellcode", IFN.genes))) -> IFN.gene.expression

IFN.gene.expression
```

I've selected a number of genes of interest, picked out of the leading edge of the GSEA.

```{r select_genes}
select.genes <- c("Ifitm3", "B2m", "Bst2", "Stat1", "Cxcl10", "Psmb9", "Psmb8", "Psmb10", "Psme1", 
              "Isg15", "Psme2", "Tap1", "Rnf213", "Zbp1", "Rsad2", "Ifit3", "Lgals3bp", "Tapbp", "Xaf1", 
              "Gbp3", "Irf1", "Ifi35", "Herc6", "Parp12", "Nmi", "Parp14", "Rtp4", "Lap3", "Eif2ak2", "Cd74", 
              "Samd9l", "Ddx58", "Ogfr", "Psma2")
```


prep for plotting.


```{r prep}
IFN.gene.expression %>% 
  pivot_longer(!cellcode, names_to = "gene", values_to =  "umi") %>% 
  left_join(enframe(mc@mc, "cellcode", "MC")) %>% 
  na.omit() %>% 
  mutate(MC = as.factor(MC)) %>% 
  filter(gene %in% select.genes) -> violin.data

violin.data
```


```{r violins, fig.width=12}
violin.data %>% 
  ggplot(aes(x = MC, y = umi, fill = MC))+
    geom_violin(scale = "width")+
    facet_wrap(~gene, scales = "free")+
    theme( strip.background = element_blank())

ggsave(here("Figs", "tumor", "MC_IFN_Violins.pdf"), width = 12, height = 6, scale = 1.2)
```


