---
title: "Compare IFN type I vs type II"
author: "Kaspar Bresser"
date: "11/10/2021"
output: 
  github_document:
    toc: true
  html_document: 
    theme: simplex
    highlight: pygments
    code_folding: show
    self_contained: TRUE
    toc: yes
    toc_float:
      collapsed: no
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      autodep  = TRUE,
                      cache = FALSE,
                      fig.width = 5,
                      fig.asp = 0.618,
                      fig.align = "center")
```

In the bulk and scRNAseq analysis of QPCTL WT and KO mice we detected a enhanced IFN response signature in tumors in a completely KO setting. In the analyses below we'll investigate if this is likely due to type I or type II IFN sensing. To figure out whether we can demonstrate this I made use of the [CytoSig](https://cytosig.ccr.cancer.gov/) database.

```{r loading}
library(here)
library(metacell)
library(tidytext)
library(rstatix)
library(GGally)
library(biomaRt)
library(tidyverse)
```

# CytoSig analysis (from dataset)

## Import and tidy

Import the cytosig dataset. Each column represents a cytokine treatment experiment on a cell model under certain conditions. Each row represents a target gene. Each value is a log2 fold change between treatment and control conditions. 


```{r import_cytosig_database}
cytosig.data <- read_tsv(here( "Data", "Cytosig", "diff.merge"))

cytosig.data
```

Select only to columns that relate to our cytokines of interest, switch to long form, and extract cytokine identity into a separate column. 

```{r tidy_data}
cytosig.data %>% 
  select(Gene, contains("IFNA"), contains("IFNB"), contains("IFNG")) %>% 
  pivot_longer(cols = -Gene, names_to = "assay", values_to = "FC") %>% 
  mutate(cytokine = str_extract(assay, "^IFN\\w{1}")) -> cytosig.IFN

cytosig.IFN
```

## Calculate scores

Remove any `NA`'s that were formed, as some genes were probably not detected in each assay. Then calculate the median log2 FC. Probably not the ideal metric, but should give us a rough idea of the association between gene-induction and IFN subtype. 

```{r caluclate_scores}
cytosig.IFN %>% 
  na.omit() %>% 
  group_by(Gene, cytokine) %>% 
  summarise(score = median(FC, na.rm = T)) %>% 
  pivot_wider(names_from = cytokine, values_from = score) -> cytosig.IFN

cytosig.IFN
```


Lets have a look at the correlations between these signatures. 

```{r correlations}
ggpairs(cytosig.IFN, columns = c("IFNA", "IFNB", "IFNG"))
```

## Map mouse to human gene symbols

The CytoSig database was constructed based on human data, our data is a mouse data-set, can use `biomaRt` to map these to each other, af far as possible. 

Retrieve the mouse gene symbols. 

```{r get_symbols, cache=TRUE}
human <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
mouse <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")

conversion <- getLDS(attributes = c("hgnc_symbol"), 
                     filters = "hgnc_symbol", 
                     values = cytosig.IFN$Gene , 
                     mart = human, 
                     attributesL = c("mgi_symbol"), 
                     martL = mouse, 
                     uniqueRows=T)

tail(conversion, 10)
```


Now import the tumor data.

```{r import_tumor}
scdb_init(here( "Data", "Metacell_files_tumor"), force_reinit=T)
mc.tumor <- scdb_mc("QPCTL_Tum_MC2")

lfp <- as_tibble(log2(mc.tumor@mc_fp), rownames = "Gene" )
```

MetaCell 12 was the cell cluster that exhibited an enhanced IFN signature.

Join with conversion table and with cytosig data.

```{r combine_sets}
lfp %>% 
  rename(MC12 = `12`) %>% 
  dplyr::select(Gene, MC12) %>% 
  inner_join(conversion, c("Gene" = "MGI.symbol")) %>% 
  inner_join(cytosig.IFN, c("HGNC.symbol" = "Gene")) -> combined.data
```

## Compare signatures

Check the correlations.

```{r correlations2}
ggpairs(combined.data, columns = c("IFNA", "IFNB", "IFNG", "MC12"), )
```


Gene-enrichment in MC12 does not seem to have a stronger correlation with any of the signatures. 

Another way of looking at the data could be to zoom in on the top 100 enriched genes in MC12 and plot the score distribution for each of the signatures.

```{r plot_sig, fig.asp=.8}
combined.data %>% 
  slice_max(order_by = MC12, n = 100) %>% 
  pivot_longer(cols = contains("IFN"), names_to = "type", values_to = "score") %>% 
  t_test(score~type)

combined.data %>% 
  slice_max(order_by = MC12, n = 100) %>% 
  pivot_longer(cols = contains("IFN"), names_to = "type", values_to = "score") %>% 
  ggplot(aes(type, score, fill = type))+
  geom_violin(scale = "width")+
  geom_boxplot(color = "black", width = .1, fill = NA)
```


# CytoSig analysis (using webtool)

Another implementation of CytoSig, is providing the webtool with gene-expression data. The tool will attempt to find gene-expression patterns that match specific cytokine profiles in their dataset. 

To perform this analysis, I'll first write out the gene-enrichment values calculated by MetaCell. Should be able to use this as input for CytoSig. Again, CytoSig is build on human data, so I'll make the conversion and provide human gene-symbols.  

```{r write_lfp}
lfp %>% 
  inner_join(conversion, c("Gene" = "MGI.symbol")) %>% 
  select(HGNC.symbol, matches("\\d")) %>% 
  write_tsv( here( "Output", "lfp_for_cytosig.tsv"))
```

Used the file generated above as input for [CytoSig](https://cytosig.ccr.cancer.gov/), downloaded the results files and stored them in `Data/Cytosig`.

Import the results

```{r import_cytosig}
cytosig.files <- list.files(here( "Data", "Cytosig"), pattern = "Cytosig")

here( "Data", "Cytosig", cytosig.files) %>% 
  map(read_csv) %>% 
  set_names(str_extract(cytosig.files, "MC\\d+")) %>% 
  enframe("MetaCell", "data") %>% 
  unnest(cols = data) -> cytosig.results

cytosig.results
```

Lets look at the results from all MetaCells, using a stringent P value cutoff. 

```{r plot_cytosig, fig.width=12, fig.asp=1}
mc.order <- paste0("MC", 1:12)

cytosig.results %>% 
  filter(Pvalue < 0.001) %>% 
  mutate(ID = reorder_within(ID, Coef, MetaCell),
         MetaCell = factor(MetaCell, levels = mc.order)) %>% 
  ggplot(aes(ID, Coef, fill = Coef > 0))+
  geom_bar(stat = "identity", color = "black")+
  facet_wrap(~MetaCell, scales = "free_y")+
  scale_x_reordered()+
  coord_flip()

ggsave(here( "Figs", "tumor", "cytosig_all.pdf"), height = 10, width = 12)
```

Or focus on the MetaCell of interest, MC12, plotting all results. 


```{r plot_MC12, fig.asp=.6, fig.width=8}
cytosig.results %>% 
  filter(MetaCell == "MC12") %>% 
  filter(Pvalue < 0.05) %>% 
  mutate(ID = reorder(ID, -Coef)) %>% 
    ggplot(aes(ID, Coef, fill = Coef > 0))+
    geom_bar(stat = "identity", color = "black")+ 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggsave(here( "Figs", "tumor", "cytosig_MC12.pdf"), height = 4, width = 6, scale = 1.5)
```

