---
title: "Analysis flow cytometry data - spleens at steady-state"
author: "Kaspar Bresser"
date: "20/10/2021"
output: 
  github_document:
    toc: true
  html_document: 
    theme: simplex
    highlight: pygments
#    code_folding: show
    self_contained: TRUE
    toc: yes
    toc_float:
      collapsed: no
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      autodep  = TRUE,
                      cache = FALSE,
                      fig.width = 5,
                      fig.asp = 0.618,
                      fig.align = "center")
```

Below the unbiased flow analysis of spleen samples from QPCTL-WT and QPCTL-KO mice. 

Import the necessary packages


```{r loading, echo=FALSE}
library(flowCore)
library(flowSpecs)
library(tidyverse)
library(umap)
library(ggpubr)
library(rstatix)
library(here)
```


## Import and prep data

Read in the fcs files using the [flowCore](https://bioconductor.org/packages/release/bioc/html/flowCore.html) package.

```{r import_fcs, warning=FALSE}
fset <- read.flowSet( path = here("Data", "spleen") )

fset
```

Down-sample to 1000 cells from each file.

```{r downsample_files}
set.seed(667)

dsFilt <- sampleFilter(size = 1000, filterId="dsFilter")
filter.result <- flowCore::filter(fset, dsFilt)
fset <- Subset(fset, filter.result)
```

We'll perform a logicle transformation to the fluorescence channels. We can use 1 sample to calculate the transformation, and then apply it to all the rest. 

```{r transform}
toTrans <- as.character(grep(pattern = "FJ", x = colnames(fset), value = T ))
toTrans

lgcl <- estimateLogicle(fset@frames$`export_20191014 Phenotyping QPCTL KO_M8_Spleen_CD45+.fcs`, channels =  toTrans)
trans.fset <- transform(fset, lgcl)
```

Finally, extract to a long dataframe. Use idInfo argument extracts shorthand sample name from fcs file name, see documentation for details.
Also, add a cell_id column.

```{r make_dataframe}
trans.fset %>%  
  flowSet2LongDf( idInfo = list("Sample" = "export_20191014 Phenotyping QPCTL KO_|_Spleen_CD45\\+\\.fcs")) %>% 
  rownames_to_column("cell_id") %>% 
  as_tibble() -> flow.data

flow.data
```

And save dataframe for later use.

```{r write_data}
write_tsv(flow.data, here("Output", "spleen_data.tsv"))
```

## UMAP

Lets subset the data for the markers that we want to use for the UMAP.

First get the markers available in the dataset


```{r get_marks}
trans.fset@frames$`export_20191014 Phenotyping QPCTL KO_M10_Spleen_CD45+.fcs`@parameters@data %>% 
  na.omit() %>% 
  pull(desc) -> markers

markers
```

Then select markers, drop other columns, and rename according to proteins

```{r select_marks}
flow.data %>% 
  dplyr::select(starts_with("FJ"), Sample, cell_id ) %>% 
  dplyr::select(-c(5,11)) %>% # drop channels that were not used
  setNames(., c(markers, "Sample", "cell_id")) %>% # rename columns 
  dplyr::select(!one_of("CD45", "7AAD")) -> flow.data # drop irrelevant markers

flow.data
```

Now we can perform the UMAP, and extract the coordinates as a dataframe. 

```{r umap}
set.seed(667)

reducU <- umap::umap(flow.data[1:11])

umap_plot_df <- as_tibble(data.frame(reducU$layout))

umap_plot_df
```

Now use those coordinates to perform hierarchical clustering. 

```{r clustering}
hc.norm = hclust(dist(umap_plot_df), method = "ward.D")
```

Add a 'cluster' column to the flow data, cutting the tree at 10 clusters. 

```{r add_clusters}
flow.data$cluster = factor(cutree(hc.norm, 10))
```

Add the coordinates to the flow data and plot the clusters

```{r plot_UMAP, fig.asp=.8}
flow.data %>% 
  bind_cols(umap_plot_df) %>% 
    ggplot( aes( x = X1, y = X2, color = cluster)) +
      geom_point(size = 0.2) + 
      guides(color = guide_legend(override.aes = list(size = 4)))+
      theme_classic()+
      labs(x = "UMAP1", y = "UMAP2")

ggsave( here("Figs", "Spleen_UMAP_clust.pdf"), width = 4.5, height = 4)
```

Plot the marker expression per cluster. 

```{r plot_marks, fig.width=14, fig.asp=.3}
flow.data %>% 
  gather("marker", "FI", -c(Sample, cluster, cell_id)) %>% 
    ggplot( aes(x = as.factor(cluster), y = FI, fill = as.factor(cluster)))+
      geom_violin()+
      facet_wrap(~marker, scales = "free_y", nrow = 2)+
      theme_classic()

ggsave(here("Figs", "Spleen_Violin_phenotypes.pdf"), width = 160, height = 40, units = "mm", scale = 2.5)

flow.data %>% 
  bind_cols(umap_plot_df) %>%
  gather("marker", "FI", -c(Sample, cluster, cell_id, X1, X2)) %>% 
  rename(UMAP1 = "X1", UMAP2 = "X2") %>% 
  write_tsv(here("Output", "Spleen_marker_expression_clusters.tsv"))
```


## Compare cluster frequencies

Next we can compare the cell counts of each sample to see if any phenotypes are enriched in WT or KO mice. 

Start off with counting cells in each cluster and add a genotype column.


```{r count_cells}
## Calculate statistics for the WT and KO cells in each cluster
flow.data %>% 
  count(cluster, Sample ) %>% # count amount of cells from each sample per cluster
  mutate(genotype = factor(case_when(Sample %in% c('M4', 'M5', 'M6_spleen_CD45+.fcs', 'M8', 'M10') ~ "WT",
                                     TRUE ~ "KO"), levels = c('WT', 'KO') ) ) -> cell.counts

cell.counts
```


Then calculate statistics. Perform a T test and add bonferroni correction


```{r check_significance}
cell.counts %>% 
  group_by(cluster) %>% # we want to test per cluster
  t_test(n ~ genotype ) %>%
  adjust_pvalue(method = "bonferroni") %>%
  add_significance("p.adj") %>% 
  add_xy_position(fun = "mean_se", x = "cluster", dodge = 0.8) %>% # x-position has to dodge around cluster
  mutate(p.adj = round(p.adj, 4))  -> stat.test

stat.test
```

And visualize as bar chart. 

```{r plot_counts, fig.width=8, fig.asp=.5}
ggplot(cell.counts, aes(x = cluster, y = n))+ 
  geom_bar(position = "dodge", stat = "summary", fun = "mean", aes(fill = genotype))+
  stat_summary(color = "black", geom = "errorbar", fun.data = "mean_se", 
               size = 0.4,width = 0.4, position = position_dodge(.9), 
               aes(group = genotype) )+
  stat_pvalue_manual(data = stat.test,  label = "p", 
                     tip.length = 0.01,hide.ns = F, label.size = 4 )+
  scale_fill_manual(values=c("#1B76BD", "#A51E23"))+
  theme_classic()

ggsave(here("Figs", "Spleen_barchart_counts.pdf"), width = 7, height = 4)

write_tsv(cell.counts, here( "Output", "Spleen_cell_counts.tsv"))
```





