---
title: "General analyses CAFs"
author: "Kaspar Bresser"
date: "20/10/2021"
output: 
  github_document:
    toc: true
  html_document: 
    theme: simplex
    highlight: pygments
#    code_folding: show
    self_contained: TRUE
    toc: yes
    toc_float:
      collapsed: no
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      autodep  = TRUE,
                      cache = FALSE,
                      fig.width = 5,
                      fig.asp = 0.618,
                      fig.align = "center")
```

Below some general visualizations of the tumor cell compartment of the scRNAseq data. 

```{r loading, echo=FALSE}
library(here)
library(Seurat)
library(metacell)
library(tidyverse)
```


## MetaCell 2D projection



Import the MetaCell object and the 2D coordinates.


```{r import_mc}
scdb_init(here("Data", "Metacell_files_fibroblast"))
mc <- scdb_mc("QPCTL_fibro_MC")

str(mc)
```


```{r import_coords}
MC.graph <- scdb_mc2d("QPCTL_fibro_MC")

coords <- tibble( cellcode = names(MC.graph@sc_y),
                  sc.x = MC.graph@sc_x,
                  sc.y = MC.graph@sc_y)
coords
```

### Plot MetaCells

Add the MetaCell identities to the coordinate data, and plot the 2D projection.

```{r plot_2d, fig.asp=.6}
str(mc@mc)

mc@mc %>% 
  enframe(name = "cellcode", value = "MetaCell") %>% 
  inner_join(coords) %>% 
  mutate(MetaCell = as.factor(MetaCell)) %>% 
    ggplot( aes(x = sc.x, y = sc.y, color = MetaCell))+ 
      geom_point(size = 1.5)+
      theme_classic()+ 
      guides(color = guide_legend(override.aes = list(size = 5)))

ggsave(filename = here("Figs", "caf", "general_2d_proj_MCs.pdf"), device = "pdf", width = 4,height = 3.2)
```


### Plot gene-expression

get gene-expression of select genes

```{r get_expression}
genes <- c('Acta2', 'Col14a1', 'Inhba', 'Tagln', 'C3', 'Gsn', 
           'Saa3', "Mt1", "Tgfb1", "Ccl2", "Cxcl5")


seurat.obj <- read_rds(here("Data", "seurat_Fibr.rds"))
seurat.obj <- NormalizeData(seurat.obj, assay = "RNA", normalization.method = "CLR")


seurat.obj %>% 
  GetAssayData( slot = "data", assay = "RNA")%>% 
#  as.matrix() %>%
  t() %>% 
  as.data.frame() %>% 
  as_tibble(rownames = "cellcode") %>% 
  select(one_of(c("cellcode", genes))) -> expression.data

expression.data
```

Join with coordinates.

```{r prep_for_plot}
expression.data %>% 
  right_join(coords) %>% 
  pivot_longer(cols = -c("sc.y", "sc.x", "cellcode"), names_to = "gene", values_to = "norm_umi") -> for.plot
```


Cap the normalized UMI counts at `2` to make a nicer visualization, and plot.

```{r 2d_genes, fig.width=8, fig.asp=.8}
for.plot %>% 
  mutate(norm_umi = ifelse(norm_umi > 1.5, 1.5, norm_umi)) %>% 
  group_by(gene) %>% 
  arrange(norm_umi) %>% 
    ggplot( aes(x=sc.x, y=sc.y, color = norm_umi))+ 
      geom_point(size = 0.3)+
      scale_color_gradient2(low= "dodgerblue3", high= "red3", mid= "white", midpoint = 0)+
      theme_classic()+
      facet_wrap(~gene)

ggsave(filename = here("Figs", "caf", "general_2d_proj_genes.pdf"), width = 8, height = 7, useDingbats = F)
```

## Cell counts in experiment

Plot the number of cells in each MetaCell. 


```{r cell_counts, }
mc@mc %>% 
  enframe(name = "cellcode", value = "MetaCell") %>% 
  mutate(MetaCell = as.factor(MetaCell)) %>% 
    ggplot(aes(x = MetaCell, fill = MetaCell))+
      geom_bar(stat = "count")+
      theme_classic()+
      theme(legend.position = "none")

ggsave(filename = here("Figs", "caf", "general_cell_counts.pdf"), width = 4,height = 2, scale = 1.2)
```


## MetaCell composition

We profiled 3 QPCTL-KO TMEs and 3 QPCTL-WT TMEs. Plot the sample composition of each MetaCell. The hashtag info is in the mat object.

```{r get_hashtags}
mat.obj <- scdb_mat("QPCTL_exp_Fibr")

mat.obj@cell_metadata %>% 
  select(new_hash.ID) %>% 
  as_tibble(rownames = "cellcode") %>% 
  filter(str_detect(new_hash.ID, "HTO")) %>% 
  mutate(sample = fct_recode(new_hash.ID, WT_1 = "HTO1", WT_2 = "HTO2", WT_3 = "HTO3",
                                          KO_1 = "HTO4", KO_2 = "HTO5", KO_3 = "HTO6")) -> hashtags

hashtags
```

Combine with MetaCells and normalize within samples. 

```{r calc_composition}
mc@mc %>% 
  enframe(name = "cellcode", value = "MetaCell") %>% 
  inner_join(hashtags) %>% 
  count(MetaCell, sample) %>%
  group_by(sample)%>%
  mutate(normalized.count = (n/sum(n))*1000 ) %>% 
  mutate(MetaCell = as.factor(MetaCell) ) -> sample.counts

sample.counts
```

And plot

```{r plot_composition}
ggplot(sample.counts, aes(fill = sample, y=normalized.count, x=MetaCell)) + 
  geom_bar( stat="identity", position = "fill")+
  scale_fill_manual(values=c("#2F4F4F", "#528B8B", "#B4CDCD", "#CD5555","#A52A2A", "#8B2323"))+
  theme(legend.title = element_blank())

ggsave(filename = here("Figs", "caf", "general_sample_compostition.pdf"), width = 4, height = 3, useDingbats=FALSE)
```



