---
title: "Marker genes CAFs"
author: "Kaspar Bresser"
date: "22/10/2021"
output: 
  github_document:
    toc: true
  html_document: 
    theme: simplex
    highlight: pygments
    code_folding: show
    self_contained: TRUE
    toc: yes
    toc_float:
      collapsed: no
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      autodep  = TRUE,
                      cache = FALSE,
                      fig.width = 5,
                      fig.asp = 0.618,
                      fig.align = "center")
```



```{r loading, echo=FALSE}
library(here)
library(Seurat)
library(metacell)
library(tidyverse)

```


# Import and normalize

Import the Seurat object containing the CAFs.

```{r import_seurat}
scdb_init(here("Data", "Metacell_files_fibroblast"), force_reinit = T)
mc <- scdb_mc("QPCTL_fibro_MC")

(seurat.object <- read_rds(here("Data", "seurat_Fibr.rds")))
```

Subset on only those cells that were assigned to a MetaCell, add MetaCell identity as metadata. 

```{r subset_data}
seurat.object <- subset(seurat.object, cells = names(mc@mc))
seurat.object <- AddMetaData(seurat.object, mc@mc, "MetaCell")

seurat.object@meta.data %>% 
  mutate( genotype = case_when(hash.ID %in% c("HTO1", "HTO2", "HTO3") ~ "WT",
                              TRUE ~ "KO")) -> seurat.object@meta.data

seurat.object
```

Normalize the RNA data.


```{r normalize_seurat, cache=TRUE, cache.lazy = FALSE}
seurat.object <- NormalizeData(seurat.object, normalization.method = "CLR", assay = "RNA")
```

Next, import the antibody capture data to investigate protein expression levels. The seurat object contains all ADT data, so subset for the fibroblast cells, add that to the `seurat.object`, and normalize. 

```{r add_ADT}
seurat.ADT <- read_rds(here("Data", "seurat_adt.rds"))
seurat.ADT <- subset(seurat.ADT,  cells = names(mc@mc))

seurat.object[["antibody_capture"]] <- seurat.ADT[["antibody_capture"]]

seurat.object <- NormalizeData(seurat.object, normalization.method = "CLR", assay = "antibody_capture")
```

# Plot CAF markers

Will look at a number of genes that are of interest to CAF biology. Used the [Nature Communations paper from Grauel and collegues](https://www.nature.com/articles/s41467-020-19920-5) as a reference. 

C3 and Cd34 are described as iCAF markers. iCAFs are known to be Il1b driven, so I'll look at expression of Il1r2. Furthermore, [Dcn](https://www.oncotarget.com/article/23869/text/) and [Ltbp1](https://www.uniprot.org/uniprot/Q14766) are known to bind and inhibit function of TGFb.


Plot expression in all MetaCells

```{r markers_all, fig.width=10}
Idents(seurat.object) <- "MetaCell"

VlnPlot(seurat.object, 
        features = c("Acta2", "Itgb1", "C3", "Il1r2", "Ltbp1", 
                     "Tagln", "Col12a1", "Cd34", "Ly6c1", "Dcn"), pt.size = 0.5 , ncol = 5)

ggsave(here("Figs", "caf", "markers_violin.pdf"), width = 3.5, height = 1.5, scale = 3)
```




Or zoom in on the iCAF (MC4) and myCAF (MC2) MetaCells

```{r markers_less, fig.width=8}
VlnPlot(seurat.object, 
        features = c("Acta2", "Itgb1", "C3", "Il1r2", "Ltbp1", 
                     "Tagln", "Col12a1", "Cd34", "Ly6c1", "Dcn"), pt.size = 0.6 , idents = c("2", "4"), ncol = 5)

ggsave(here("Figs", "caf", "markers_violin_2and4_only.pdf"), width = 2.5, height = 1.5, scale = 3)
```


As an extra layer of information, we can split by the origin of the cells (KO or WT TME). 


```{r markers_less2, fig.width=8.5}
VlnPlot(seurat.object, 
        features = c("Acta2", "Itgb1", "C3", "Il1r2", "Ltbp1", 
                     "Tagln", "Col12a1", "Cd34", "Ly6c1", "Dcn"), 
        idents = c("2", "4"), split.by = "genotype", 
        pt.size = 0.6, ncol = 5 ) + theme(legend.position = "right")

ggsave(here("Figs", "caf", "markers_violin_2and4_only_split.pdf"), width = 2.5, height = 1.5, scale = 3)

seurat.object@meta.data %>% 
  count(genotype, MetaCell)
```


There seems to be some level of within MetaCell heterogeneity here. Specifically the myofibroblastic markers of MC2 appear to be mainly expressed by cells from the WT setting. 


We also used CITEseq antibodies, lets zoom in on iCAF and myCAF markers Ly6C and ITGB1.


```{r plot_ADT_less , fig.width=7}
VlnPlot(object = seurat.object, features =c("Ly6C", "ITGB1"), idents = c("2", "4"), assay = "antibody_capture", pt.size = 0.2, y.max = 3)

ggsave(here("Figs", "caf", "markers_violin_CITE.pdf"), width = 2, height = 2, scale = 1.5)
```



There's not a whole lot of cells, but we can additionally split these cells based on their origin from KO or WT TME's. 


```{r plot_ADT_less2 , fig.width=7}
VlnPlot(object = seurat.object, features =c("Ly6C", "ITGB1"), idents = c("2", "4"), 
        assay = "antibody_capture", split.by = "genotype", pt.size = 0.2, y.max = 3) + theme(legend.position = "right")

ggsave(here("Figs", "caf", "markers_violin_CITE_split.pdf"), width = 3, height = 2, scale = 1.5)
```


Interestingly it seems that KO cells that reside in MetaCell 2 express higher levels of Ly6C, and lower levels of ITGB1. 

Quick check to see if RNA follows the same pattern. 


```{r plot_ADT_less2_rna , fig.width=7}
VlnPlot(object = seurat.object, features =c("Ly6c1", "Itgb1"), idents = c("2", "4"), 
        assay = "RNA", split.by = "genotype", pt.size = 0.3) + theme(legend.position = "right")
```


